Model Finding for Recursive Functions in SMT

### 2016-01-26 ijcar2016/body.tex
Minor fix

@@

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{I}}
\newcommand{\J}{\mathcalx{J}}
\newcommand{\conv}{\mathcalx{A}}

@@

\hyphenation{counter-example counter-examples Isa-belle data-type data-types
  co-data-type co-data-types}

@@

\begin{abstract}
\noindent
SMT solvers have recently been extended with techniques for finding models
in presence of universally quantified formulas in some restricted fragments.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including well-founded (terminating) functions,
to universally quantified formulas for which these
techniques are applicable.
An empirical evaluation confirms that the approach improves 
the performance of existing solvers on benchmarks from three sources. 
The translation is implemented as a preprocessor in the CVC4 solver
and in a new higher-order model finder called Nunchaku.
\end{abstract}

@@

Many solvers based on SMT (satisfiability modulo theories) can reason about
quantified formulas using incomplete instantiation-based methods
\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods work well in the context of proving (i.e., showing
unsatisfiability), but they are of little help for finding models (i.e.,
showing satisfiability). Often, a single universal quantifier in one of the
axioms of a problem is enough to prevent the discovery of models.

In the past few years, techniques have been developed to find models for
quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} introduced a complete instantiation-based
procedure for formulas in the \relax{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
%ct occurrences of 
variables occur as direct subterms of uninterpreted
functions---e.g., $\forall x%\typ{\typeint} --- the type is not important, right?
.\;\, \con{f}( x )
\teq \con{g}( x ) + 5$.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
Other syntactic criteria extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicates. Subsequently, Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} introduced techniques for finding finite
models for quantified
formulas over uninterpreted types and types having a fixed finite
interpretation. %, such as fixed-width bit vectors and enumerated datatypes.
These techniques can
find a model for a formula such as $\forall x,\, y : \tau.\;\, x \teq
y \vee \allowbreak \neg \con{f}( x ) \teq \con{f}( y )$, where $\tau$ is an uninterpreted type.
% where the uninterpreted type $\tau$ is
% interpreted as a finite set and $\con{f}$ as injective finite map.

Unfortunately, none of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions. The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used inside a complex term on the right-hand side,
whereas the finite model finding techniques %of Reynolds et al.\
are not applicable for functions over infinite domains such as the integers or
algebraic datatypes. A simple example where both approaches fail is
$$\forall x : {\typeint}.\;\,
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$$ %.
This state of affairs is unsatisfactory, given the frequency of
recursive definitions in practice and the %impending
addition of a %dedicated
command for introducing them, \texttt{define-fun-rec},
to the SMT-LIB standard \cite{smtlib25}.

%In this paper,
We present a method for translating formulas involving recursive function
definitions into formulas where finite model finding techniques can be applied.
The recursive functions must meet a semantic criterion to be admissible
(Section~\ref{sec:prelim}). This criterion is met by well-founded (terminating)
recursive function definitions. It is not met by inconsistent
definitions such as \,$\forall x : {\typeint}.\;\, \con{f}(x) \teq \con{f}(x) +
1$.

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). A
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq \mathit{rhs}$
is translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq \mathit{rhs}[\farg{\con{f}}(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} :
\fargtype{\tau} \to \tau$ converts the abstract type to the concrete
type. Additional constraints ensure that the abstract values that are relevant
to the formula's satisfiability
exist. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and makes finite model finding possible for problems in this
class.
%\begin{conf} Detailed proofs of correctness are included in a technical
%report \cite{our-report}.\end{conf}

The approach is implemented as a preprocessor in \cvc and in the Nunchaku
model finder (Section~\ref{sec:implementations}).
%As a case study, we employed
%Nunchaku on Isabelle/HOL~\cite{nipkow-et-al-2002} formalizations of
%coinductive lists and binary trees (Section~\ref{sec:case-studies}).
%
We evaluated the two evaluation on benchmarks from
IsaPlanner~\cite{DBLP:conf/itp/JohanssonDB10}, Leon~\cite{blanc2013overview},
and Isabelle/HOL, to demonstrate that this translation improves the
effectiveness of the SMT solvers \cvc and \ziii for finding countermodels to
verification conditions (Section~\ref{sec:evaluation}). Unlike earlier work,
our approach can be combined with off-the-shelf SMT solvers
(Section~\ref{sec:related-work}).

@@

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an \innerkw{if}--\innerkw{then}--\innerkw{else} function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
We fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$}
for each $\tau \in \stypes{\Sigma}$ and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
$\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote %ct a metavariable ranging over 
terms of type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{rep}When applied to terms, the symbol $=$ denotes syntactic equality.\end{rep}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I,$
each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ to an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{rep};
otherwise, it is \emph{$T$-unsatisfiable}\end{rep}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent} 
if each $T$-entails the other.
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature 
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where 
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not in $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$ 
with uninterpreted symbols} constructed as above.

Unconventionally,
we consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
uninterpreted. Their
semantics is the same as for standard quantified formulas $\forall \vec x.\; \varphi$.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1 : \tau_1$, $\ldots,$ $x_n : \tau_n$
and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
%I don't think it makes much sense to have any quantifiers within function bodies, 
%since they aren't operational, that's why I put quantifier-free. --ajr
%%\jb{Actually, in Isabelle we can have them and we use them sometimes. It certainly
%% makes sense. I can give you real examples if you like.}
We write $\exists
\vec x.\; \varphi$ as an abbreviation for $\neg \forall \vec x.\; \neg \varphi$.

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \ldots, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
$(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi$,
%
where $\con{f}_1, \ldots, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \ldots, n$,
%
%\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
%\jb{do we? I thought that the admissibility criterion below is enough.}
% should be enough, unless I'm missing something? -ajr
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{goal} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a ground formula $\psi$ 
%\ct{I suppose $\psi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if
\[\psi \models_T \smash{\bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}}\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\psi$ closed under function expansion with respect to $\DDD$,
the formula
$\psi \wedge \bigwedge \Delta$ is also $T$-satisfiable.
%\ct{I}
\end{definition}

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying the translation described in
Section~\ref{sec:encoding}. It is useful to connect it to the
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, 
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$,
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \typeint$, can be shown well founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

\begin{theorem}\label{thm:adm}
If $\DDD$ is a set of well-founded function definitions for\/
$\sfundefs{\Sigma} = \{\con{f}_1,\ldots,\con{f}_n\}$, then it is admissible.
\end{theorem}
\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\I_0$ be the restriction of $\I$ to the function symbols in $\sfuns{\Sigma} -
\sfundefs{\Sigma}$. Because well-founded definitions uniquely characterize
the interpretation of the functions they define, there exists a
$\Sigma$-interpretation $\J$ that extends $\I_0$ such that $\J \models \Delta$.
%
Since $\psi$ is closed under function expansion, it already constrains the
functions in $\sfundefs{\Sigma}$ recursively as far as is
necessary for interpreting $\psi$. Thus, any point $v$ for which
$\const{f}_i^\I(v)$ is needed for interpreting $\psi$ will have its
expected value according to its definition and hence coincide with $\J$.
And since $\psi^\I$ does not depend on the interpretation at the other
points, $\J$ is, like $\I$, a model of $\psi$.
Since $\J \models \Delta$ by assumption, we have $\J \models \psi \wedge
\bigwedge \Delta$ as desired.
%We have
%$\varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \}$.
\qed
\end{proofsketch}

@@

Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.
%We conjecture that a large fragment of mixed recursive--corecursive
%specifications are

@@

and $\absconstraints( L_0 )$ consists of the single constraint

@@

since it entails $\varphi_\con{s}[(\con{c}-i)/x]$ for $i = 0, \ldots, n$,

>>>>

%!TEX root =  pap.tex

@@

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
%CT somehow this compiles into a Roman font, removed for now
%\newcommand{\I}{\mathcalx{I}}
%\newcommand{\J}{\mathcalx{J}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{{\mathcal{I}'}}
\newcommand{\conv}{\mathcalx{A}}

@@

\hyphenation{counter-example counter-examples Isa-belle data-type data-types
  co-data-type co-data-types bench-mark bench-marks}

@@

\begin{abstract}
\noindent
SMT solvers have recently been extended with techniques for finding models
in the presence of universally quantified formulas in some restricted fragments
of first-order logic.
This paper introduces a translation which reduces axioms specifying a large
class of recursive functions, including well-founded (terminating) functions,
to universally quantified formulas for which these
techniques are applicable.
An empirical evaluation confirms that the approach improves
the performance of existing solvers on benchmarks from three sources.
The translation is implemented as a preprocessor in the CVC4 solver
and in a new higher-order model finder called Nunchaku.
\end{abstract}

@@

Many solvers based on SMT (satisfiability modulo theories) can reason about
quantified formulas using incomplete instantiation-based methods
\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods work well for proving the
unsatisfiability of an input set of formulas, 
but they are of little help for finding models of them
when they are satisfiable. Often, a single universal quantifier in one of the
axioms of a problem is enough to prevent the discovery of models.

In the past few years, a few techniques have been developed to find models for
quantified formulas in SMT.
Ge and de Moura \cite{GeDeM-CAV-09} introduced a complete instantiation-based
procedure for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all
%ct occurrences of
variables occur as direct subterms of uninterpreted
functions, as in $\forall x : \typeint.\; \con{f}( x ) \teq \con{g}( x ) + 5$.
% since all
% occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and
% $\con{g}$.
Other syntactic criteria extend
this fragment slightly, including cases when variables occur as arguments of
arithmetic predicate symbols. Subsequently, Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013} introduced techniques for finding finite
models for quantified
formulas over uninterpreted types and types having a fixed finite
interpretation. %, such as fixed-width bit vectors and enumerated datatypes.
These techniques can
find a model for a formula such as $\forall x,\, y : \tau.\;\, x \teq
y \vee \allowbreak \neg \con{f}( x ) \teq \con{f}( y )$, where $\tau$ is an uninterpreted type.
% where the uninterpreted type $\tau$ is
% interpreted as a finite set and $\con{f}$ as injective finite map.

Unfortunately, none of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions. The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used inside a complex term on the right-hand side
of the definition,
whereas the finite model finding techniques %of Reynolds et al.\
are not applicable for functions over infinite domains such as the integers or
algebraic datatypes. A simple example where both approaches fail is
$$\forall x : {\typeint}.\;\,
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$$ %.
where $\lite$ is the `if--then--else' operator.
This state of affairs is unsatisfactory, given the frequency of
recursive definitions in practice, underscored by the recent
addition of a %dedicated
command for introducing them, \texttt{define-fun-rec},
to the SMT-LIB standard \cite{smtlib25}.

%In this paper,
We present a method for translating formulas involving recursive function
definitions into formulas where finite model finding techniques can be applied.
The definitions must meet a semantic criterion to be admissible 
(Section~\ref{sec:prelim}). 
This criterion is general enough to include well-founded (terminating)
recursive function definitions and restrictive enough to exclude inconsistent
equations such as \,$\forall x : {\typeint}.\;\, \con{f}(x) \teq \con{f}(x) +
1$.

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). A
recursive equation $\forall x : \tau.\;\, \const{f}(x) \teq t$
is translated to
$\forall a :\nobreak
\fargtype{\tau}.\allowbreak\;\, \const{f}(\farg{\con{f}}(a)) \teq t[\farg{\con{f}}(a)/x]$, where
%%% To me, "abstract" and "concrete" are just convenient labels to refer to the
%%% thing. I don't want readers to start wondering, "What is an abstract type?"
%%% So I find it important to leave "uninterpreted" outside of parentheses. As
%%% for italics, I went for quotes instead, since this is an informal explanation
%%% anyway. Later, we define \emph{abstract type} with italics. --JB
$\fargtype{\tau}$ is an ``abstract'' uninterpreted type and $\farg{\con{f}} :
\fargtype{\tau} \to \tau$ is an uninterpreted function
from $\fargtype{\tau}$ to the corresponding concrete type $\tau$. 
Additional constraints ensure that the abstract values that are relevant
to the formula's satisfiability
exist. The translation preserves satisfiability and, for admissible definitions,
unsatisfiability, and makes finite model finding possible for problems in this
class.
%\begin{conf} Detailed proofs of correctness are included in a technical
%report \cite{our-report}.\end{conf}

The approach is implemented as a preprocessor in \cvc and in the Nunchaku
model finder (Section~\ref{sec:implementations}).
%As a case study, we employed
%Nunchaku on Isabelle/HOL~\cite{nipkow-et-al-2002} formalizations of
%coinductive lists and binary trees (Section~\ref{sec:case-studies}).
%
We evaluated the two implementations on benchmarks from
IsaPlanner~\cite{DBLP:conf/itp/JohanssonDB10}, Leon~\cite{blanc2013overview},
and Isabelle/HOL, to demonstrate that this translation improves the
effectiveness of the SMT solvers \cvc and \ziii in finding countermodels to
verification conditions (Section~\ref{sec:evaluation}). Unlike earlier work,
our approach relies on off-the-shelf SMT solvers
(Section~\ref{sec:related-work}).

@@

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an \innerkw{if}--\innerkw{then}--\innerkw{else} function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
We fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$}
for each $\tau \in \stypes{\Sigma}$ and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
(Well-typed) $\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote %ct a metavariable ranging over
terms of type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, a variable tuple $\vec x = (x_1^{\tau_1},\dotsc,x_n^{\tau_n})$ and 
a term tuple $\vec t = (t_{\smash{1}\vphantom{n}}^{\tau_1},\dotsc,t_n^{\tau_n})$, 
we write $u[\vec t/\vec x]$ to denote the result of simultaneously replacing
all occurrences of $x_i$ with $t_i$ in $u$, for each $i=1,\dotsc,n$.
\begin{rep}When applied to terms, the symbol $=$ denotes syntactic equality.\end{rep}

@@

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each type $\tau \in \stypes{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in~$\I,$
each function symbol $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfuns{\Sigma}$
to a total function $\con{f}^\I : \tau_{\smash{1}}^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$,
and each variable $x:\tau$ of $\svars{\Sigma}$ to an element of $\tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where
$\Sigma$ is a signature and $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$, closed under variable reassignment
(i.e., for every $I \in \Mo$, every $\Sigma$-interpretation that differs
from $I$ only on the variables of $\svars \Sigma$ is also in $\Mo$).
A $\Sigma$-formula $\varphi$ is \emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$\begin{rep};
otherwise, it is \emph{$T$-unsatisfiable}\end{rep}.
A formula $\varphi$ \emph{$T$-entails} $\psi$, written $\varphi \models_T \psi$,
if all interpretations in $\Mo$ that satisfy $\varphi$ also satisfy $\psi$.
Two formulas $\varphi$ and $\psi$ are \emph{$T$-equivalent}
if each $T$-entails the other.
If $T_1 = (\Sigma_1, \Mo_1)$ is a theory and $\Sigma_2$ is a signature
with $\sfuns{\Sigma_1} \cap \sfuns{\Sigma_2} = \emptyset$,
the \emph{extension of $T_1$ to $\Sigma_2$} is the theory $T = (\Sigma, \Mo)$ where
$\sfuns{\Sigma} = \sfuns{\Sigma_1} \cup \sfuns{\Sigma_2}$,
$\stypes{\Sigma} = \stypes{\Sigma_1} \cup \stypes{\Sigma_2}$,
and $\Mo$ is the set of all $\Sigma$-interpretations $\I$
whose $\Sigma_1$-reduct is a model of $T_1$.
We refer to the symbols of $\Sigma_2$ that are not in $\Sigma_1$ as \emph{uninterpreted}.
For the rest of the paper, \relax{we fix a theory $T = (\Sigma, \Mo)$
with uninterpreted symbols} constructed as above.

Unconventionally,
we consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
uninterpreted. Their
semantics is the same as for standard quantified formulas $\forall \vec x.\; \varphi$.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1^{\tau_1}$, $\dotsc,$ $x_n^{\tau_n}$
and $\varphi$ is a quantifier-free formula
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?}
%I don't think it makes much sense to have any quantifiers within function bodies,
%since they aren't operational, that's why I put quantifier-free. --ajr
%%\jb{Actually, in Isabelle we can have them and we use them sometimes. It certainly
%% makes sense. I can give you real examples if you like.}
We write $\exists
\vec x.\; \varphi$ as an abbreviation for $\neg \forall \vec x.\; \neg \varphi$.

\begin{definition}\rm
A formula $\varphi$ is in \emph{definitional form with respect to}
$\{ \con{f}_1, \dotsc, \con{f}_n \} \subseteq \sfuns{\Sigma}$ if it is of the
form
%
%\kern-\abovedisplayskip
%\kern+\abovedisplayshortskip
%
$(\forallf{\con{f}_1} \vec x_1.\; \varphi_1) \wedge \cdots \wedge
(\forallf{\con{f}_n} \vec x_n.\; \varphi_n) \wedge \psi$,
%
where $\con{f}_1, \dotsc, \con{f}_n$ are distinct function symbols,
%ct added
$\forallf{\con{f}_i} \vec x_i.\; \varphi_i$ is a function definition
for $i = 1, \dotsc, n$,
%
%\ct{I think we need more conditions here on how a symbol $\con{f}_i$ can occur in $\varphi_j$.}
%\jb{do we? I thought that the admissibility criterion below is enough.}
% should be enough, unless I'm missing something? -ajr
and $\psi$ contains no function definitions.
We call $\psi$ the \emph{goal} of $\varphi$.
%\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
%(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
% Now fixed above -ajr
\end{definition}

@@

\begin{definition}\rm
Given a set of function definitions
$\DDD = \{ \forallf{\con{f}_1} \vec x_1.\; \varphi_1, \dotsc, \forallf{\con{f}_n} \vec x_n.\; \varphi_n \}$,
a ground formula $\psi$
%\ct{I suppose $\psi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if
\[\psi \models_T \smash{\bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x_i ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}}\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula
$\psi$ closed under function expansion with respect to $\DDD$,
the formula
$\psi \wedge \bigwedge \Delta$ is also $T$-satisfiable.
%\ct{I}
\end{definition}

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying our translation, described in
Section~\ref{sec:encoding}. It is useful to connect it to the
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the defined function be terminating when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\,
\con{p}( x ) \teq \lite\bigl( x \leq 0,\allowbreak\;  1,\allowbreak\;
  2 * \con{p}( x - 1 ) \bigr)$,
where the theory $T$ is integer arithmetic extended with the uninterpreted symbol
$\con p:\typeint \rightarrow \typeint$, can be shown to be well founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

\begin{theorem}\label{thm:adm}
If $\DDD$ is a set of well-founded function definitions for\/
$\sfundefs{\Sigma} = \{\con{f}_1,\dotsc,\con{f}_n\}$, then it is admissible.
\end{theorem}

\begin{proofsketch}
Let $\psi$ be a satisfiable formula closed under function expansion with
respect to $\DDD$. We show that $\psi \wedge \bigwedge \Delta$ is also
satisfiable. Let $\I$ be a model of $\psi$, and
let $\I_0$ be the reduct of $\I$ to the function symbols in $\sfuns{\Sigma} \setminus
\sfundefs{\Sigma}$. Because well-founded definitions uniquely characterize
the interpretation of the functions they define, there exists a
$\Sigma$-interpretation $\J$ that extends $\I_0$ such that $\J \models \Delta$.
%
Since $\psi$ is closed under function expansion, it already constrains the
functions in $\sfundefs{\Sigma}$ recursively as far as is
necessary for interpreting $\psi$. Thus, any point $v$ for which
$\const{f}_i^\I(v)$ is needed for interpreting $\psi$ will have its
expected value according to its definition and hence coincide with $\J$.
And since $\psi^\I$ does not depend on the interpretation at the other
points, $\J$ is, like $\I$, a model of $\psi$.
Since $\J \models \Delta$ by assumption, we have $\J \models \psi \wedge
\bigwedge \Delta$ as desired.
%We have
%$\varphi \,\models_T\, \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \varphi ) \}$.
\qed
\end{proofsketch}

@@

Another useful class of function definitions is that of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. 
Intuitively, productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the function defined by
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
%%% If what follows ':' is a complete sentence, the standard Am.E. convention
%%% is to capitalize it. --JB
falls within this class: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specifies the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.
%We conjecture that a large fragment of mixed recursive--corecursive
%specifications are

@@

and $\absconstraints( L_0 )$, defined by equation (\ref{eq:xl}), consists of the single constraint

@@

since it entails $\varphi_\con{s}[(\con{c}-i)/x]$ for $i = 0, \dotsc, n$
and contains only $\con{s}$ applications corresponding to $\con{s}( \con{c}-i )$ for $i = 0, \dotsc, n$.
